---
title: "Final Project"
subtitle: "Data Exploration"
author: ""
date: "2019-Feb-25"
fontsize: 8 pt
output:
  #    beamer_presentation: #everything after hash "#" will be ignored 
pdf_document:
number_sections: false
fig_width: 6
fig_height: 4
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      #include = TRUE, 
                      eval = TRUE, 
                      
                      fig.width = 6, fig.height = 4,
                      results='hold',
                      warning = FALSE,
                      message=FALSE,
                      cache = TRUE,
                      digits = 3) 
```

```{r,echo=FALSE,results='hide', message=FALSE}
# Packages and Data Parser
#Note: Be sure plyr is loaded before dplyr
PackageList =c("plyr","dplyr","tidyverse","lubridate","censusapi","forcats")
NewPackages=PackageList[!(PackageList %in% 
                            installed.packages()[,"Package"])]
if(length(NewPackages)) install.packages(NewPackages,repos = "http://cran.us.r-project.org")
lapply(PackageList,require,character.only=TRUE)#array function

options(tibble.print_max = Inf, tibble.print_min = 20) 
set.seed(1) #Always set the seed for reproducibility
```

Load data and begin generating exploratory plots.

```{r}
source("taxiData.R")
#Load records from taxi data source and store as a dataframe
data1m <- taxiData(100000)
data1mDF <- data.frame(data1m$rawData)
rm(data1m) #Drop old data
data1mDF$dropoff_community_area<-as.factor(data1mDF$dropoff_community_area)
#Load census data by Chicago Community Area
ccaData <- read.csv("ChicagoCCAData.csv")
ccaData$CCA.Code<-as.factor(ccaData$CCA.Code)
ccaData$GEOG<-as.factor(ccaData$GEOG)
#Look up name of pickup CCA and add to df
df<- data1mDF %>% left_join(select(ccaData,GEOG,CCA.Code),by=c("pickup_community_area"="CCA.Code"))
names(df)[names(df) == 'GEOG'] <- 'pickup_area'
#Look up name of dropoff CCA and add to df
df<- df %>% left_join(select(ccaData,GEOG,CCA.Code),by=c("dropoff_community_area"="CCA.Code"))
names(df)[names(df) == 'GEOG'] <- 'dropoff_area'

```
Clean data. For instance, "Chicago Elite Cab Corp. (Chicago Carriag" and "Chicago Elite Cab Corp." are clearly the same firm. We replace the former with the latter. 
We add factors for the Start Date (as opposed to date time), End Date, and Start Day of Week.
* In company, set "Chicago Elite Cab Corp. (Chicago Carriag" equal to "Chicago Elite Cab Corp."
* Add factor for Start Date (not date time)
* Add factor for End Date (not date time)
* Add factor for start day of the week

```{r}
df$company[data1mDF$company=="Chicago Elite Cab Corp. (Chicago Carriag"] <- "Chicago Elite Cab Corp."
df$startDate<-date(data1mDF$trip_start_timestamp)
df$endDate<-date(data1mDF$trip_end_timestamp)
df$endDay<-wday(data1mDF$trip_end_timestamp,label=TRUE)
df$endMonth
```

Throw out all rows where NAs exist for dropoff/pickup location and convert NA to explicit for company (necessary in summary calcs)

```{r}
#Replace <NA> with string "NA" which can be used for grouping
#df<-df %>% drop_na(c("dropoff_area","pickup_area"))
df$company <- fct_explicit_na(df$company,"NA")
```


## Start Summary Tables

Create summary tables
* Ride count by Company
* Ride count by Dropoff Area
* Ride count by Pickup Area
* Rice count, avg fare, avg distance by pickup area
* Rice count, avg fare, avg distance by dropoffS area
```{r}
#Note: I am calling dplyr's functions directly to avoid issues with summarize
#Trips by Company
df %>% dplyr::group_by(company) %>% tally(sort=TRUE)
#Trips by Dropoff Area
df %>% dplyr::group_by(dropoff_area) %>% tally(sort=TRUE)
#Trips by Origination and Destination
df %>% dplyr::group_by(pickup_area,dropoff_area) %>% tally(sort=TRUE)
#Trips by Pickup Area
df %>% dplyr::group_by(pickup_area) %>% tally(sort=TRUE)
#Avg Fare and Distance by Pickup Area
df %>% dplyr::group_by(pickup_area) %>% dplyr::summarize(total_trips=n(),mean_fare = round(mean(fare,na.rm=TRUE),2),mean_distance=round(mean(trip_miles,na.rm=TRUE),1))
#Avg Fare and Distance by Dropoff Area
df %>% dplyr::group_by(dropoff_area) %>% dplyr::summarize(total_trips=n(),mean_fare = round(mean(fare,na.rm=TRUE),2),mean_distance=round(mean(trip_miles,na.rm=TRUE),1))
#Avg Fare and Distance by Pickup and Dropoff Areas
df %>% dplyr::group_by(pickup_area,dropoff_area) %>% dplyr::summarize(total_trips=n(),
  mean_fare = round(mean(fare,na.rm=TRUE),2),
  median_fare = round(median(fare,na.rm=TRUE),2),
  min_fare=round(min(fare,na.rm=TRUE),2),
  max_fare=round(max(fare,na.rm=TRUE),2),
  mean_distance=round(mean(trip_miles,na.rm=TRUE),1),
  median_distance=round(median(trip_miles,na.rm=TRUE),1),
  min_distance=round(min(trip_miles,na.rm=TRUE),1),
  max_distance=round(max(trip_miles,na.rm=TRUE),1)
  )  %>% arrange(desc(mean_fare))
```



Generate summary calculations by month/year
```{r}
fareByMonthYear <- df %>% dplyr::group_by(month=floor_date(df$startDate, "month")) %>%
   summarize(totalFares=sum(fare))
ridesByMonthYear <- df %>% dplyr::group_by(month=floor_date(df$startDate, "month")) %>%
   tally(sort=FALSE)
```

Plot Total Fares paid in Chicago as a functon of Month
```{r}
plot(x=fareByMonthYear$month,y=fareByMonthYear$totalFares,main="Total Cab Fares by Month/Year")
```
```{r}
plot(x=ridesByMonthYear$month,y=ridesByMonthYear$n,main="Number of Cab Rides by Month/Year")
```



```{r}

#ggplot(data = data1mDF) +
#  geom_histogram(mapping = aes(x = fare), binwidth = 5)

#ggplot(data = data1mDF) +
#  geom_bar(mapping = aes(x = company))

#ggplot(data = data1mDF) +
#  geom_bar(mapping = aes(x = company))

#ggplot(data = data1mDF, aes(x = trip_start_timestamp)) + 
# geom_density(fill = "blue")

```

```{r}
#orderedFares<-data1mDF$fare[order(data1mDF$fare,decreasing = TRUE)]

#boxplot(orderedFares[10:length(orderedFares)])
```

