---
title: "Final Project"
subtitle: "Data Exploration"
author: ""
date: "2019-Feb-25"
fontsize: 8 pt
output:
  #    beamer_presentation: #everything after hash "#" will be ignored 
pdf_document:
number_sections: false
fig_width: 6
fig_height: 4
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      #include = TRUE, 
                      eval = TRUE, 
                      
                      fig.width = 6, fig.height = 4,
                      results='hold',
                      warning = FALSE,
                      message=FALSE,
                      cache = TRUE,
                      digits = 3) 
```

```{r,echo=FALSE,results='hide', message=FALSE}
# Packages and Data Parser
#Note: Be sure plyr is loaded before dplyr
PackageList =c("plyr","dplyr","tidyverse","lubridate","censusapi","forcats")
NewPackages=PackageList[!(PackageList %in% 
                            installed.packages()[,"Package"])]
if(length(NewPackages)) install.packages(NewPackages,repos = "http://cran.us.r-project.org")
lapply(PackageList,require,character.only=TRUE)#array function

options(tibble.print_max = Inf, tibble.print_min = 20) 
set.seed(1) #Always set the seed for reproducibility
```

Load data and begin generating exploratory plots.

```{r}
source("taxiData.R")
#Load records from taxi data source and store as a dataframe
data1m <- taxiData(5000000)
df <- data.frame(data1m$rawData)
rm(data1m) #Drop old data
df$dropoff_community_area<-as.factor(df$dropoff_community_area)
df$dropoff_centroid_location.coordinates<-as.character(df$dropoff_centroid_location.coordinates)
df$pickup_centroid_location.coordinates<-as.character(df$pickup_centroid_location.coordinates)
df$trip_total<-as.numeric(df$trip_total)
#Load census data by Chicago Community Area
ccaData <- read.csv("ChicagoCCAData.csv")
ccaData$CCA.Code<-as.factor(ccaData$CCA.Code)
ccaData$GEOG<-as.factor(ccaData$GEOG)
#Look up name of pickup CCA and add to df
df<- df %>% left_join(select(ccaData,GEOG,CCA.Code),by=c("pickup_community_area"="CCA.Code"))
names(df)[names(df) == 'GEOG'] <- 'pickup_area'
#Look up name of dropoff CCA and add to df
df<- df %>% left_join(select(ccaData,GEOG,CCA.Code),by=c("dropoff_community_area"="CCA.Code"))
names(df)[names(df) == 'GEOG'] <- 'dropoff_area'
#Look up population>16yo of pickup CCA and add to df
df<- df %>% left_join(select(ccaData,POP_16OV,CCA.Code),by=c("pickup_community_area"="CCA.Code"))
names(df)[names(df) == 'POP_16OV'] <- 'Pop16Plus'
#Look up Labor Force of pickup CCA and add to df
df<- df %>% left_join(select(ccaData,IN_LBFRC,CCA.Code),by=c("pickup_community_area"="CCA.Code"))
names(df)[names(df) == 'IN_LBFRC'] <- 'IN_LBFRC'
#Look up Unemp of pickup CCA and add to df
df<- df %>% left_join(select(ccaData,UNEMP,CCA.Code),by=c("pickup_community_area"="CCA.Code"))
names(df)[names(df) == 'UNEMP'] <- 'UNEMP'
#Look up Unemp of pickup CCA and add to df
df<- df %>% left_join(select(ccaData,MEDINC,CCA.Code),by=c("pickup_community_area"="CCA.Code"))
names(df)[names(df) == 'MEDINC'] <- 'MedIncome'
```


Clean data. For instance, "Chicago Elite Cab Corp. (Chicago Carriag" and "Chicago Elite Cab Corp." are clearly the same firm. We replace the former with the latter. 
We add factors for the Start Date (as opposed to date time), End Date, and Start Day of Week.
* In company, set "Chicago Elite Cab Corp. (Chicago Carriag" equal to "Chicago Elite Cab Corp."
* Add factor for Start Date (not date time)
* Add factor for End Date (not date time)
* Add factor for End Day of week
* Add factor for End month
* Add factor for End hour

```{r}
df$company[df$company=="Chicago Elite Cab Corp. (Chicago Carriag"] <- "Chicago Elite Cab Corp."
df$startDate <- date(df$trip_start_timestamp)
df$endDate <- date(df$trip_end_timestamp)
df$endDay <- wday(df$trip_end_timestamp,label=TRUE)
df$endMonth <- month(df$trip_end_timestamp,label=TRUE)
df$endHour <- hour(str_replace(df$trip_end_timestamp,"T"," "))
```

Throw out all rows where fare, duration, or distance=0 and where NAs exist for dropoff/pickup location and convert NA to explicit for company (necessary in summary calcs)

```{r}
#Uncomment the following line to drop all records where dropoff_area and pickup_area NA
#df<-df %>% drop_na(c("dropoff_area","pickup_area"))
#Replace <NA> with string "NA" which can be used for grouping
nrowPreFilter=nrow(df)
df <- df[df$fare>0,]
df <- df[df$trip_miles>0,]
df <- df[df$trip_seconds>0,]
df <- df[df$trip_total>0,]
df$company <- fct_explicit_na(df$company,"NA")

cat("Filtering out records where fare,trip_miles,trip_seconds=0. Remaining records: ", round(100*nrow(df)/nrowPreFilter,1),"% of data. Leaves ", nrow(df)," total records.")
```
```{r}
cat("After dropping records we want to add a column for total hourly rate earned by ride: trip_total/(trip_seconds/60*60)\n")
cat("\nAdd a column for total hourly fare by ride: fare/(trip_seconds/60*60)\n")
df$totalHourlyRate=df$trip_total*60*60/(df$trip_seconds)
df$totalHourlyFare=df$fare*60*60/(df$trip_seconds)

cat("The average hourly rate across all records is $",round(mean(df$totalHourlyRate, na.rm=TRUE),2)," per hour")
cat("\nThe median hourly rate across all records is $",round(median(df$totalHourlyRate, na.rm=TRUE),2)," per hour")
cat("\n\nThe average hourly fare across all records is $",round(mean(df$totalHourlyFare, na.rm=TRUE),2)," per hour")
cat("\nThe median hourly fare across all records is $",round(median(df$totalHourlyFare, na.rm=TRUE),2)," per hour")
cat("\n\nChicago cab rates: $3.25 flag pull, $2.25/mi, $0.25 per 36 secs.")
cat("\nAssume 1 passenger:")
cat("\nFare for a 1 hour trip for a cab sitting still: $",(3.25+.25*60*60/36))
cat("\nFare for a 1 hour trip for a cab averaging 15mph: $",(3.25+.25*60*60/36+15*2.25))
cat("\nFare for a 1 hour trip for a cab averaging 30mph: $",(3.25+.25*60*60/36+30*2.25))
cat("\nFare for a 1 hour trip for a cab averaging 45mph: $",(3.25+.25*60*60/36+45*2.25))

cat("\n\nAvg cab speed:",round(mean(df$trip_miles*60*60/df$trip_seconds,na.rm=TRUE),1),"mph")

cat("\nA 1 hour trip for a cab averaging 16.5mph: $",round((3.25+.25*60*60/36+16.5*2.25),2))


```

Maps Pickups vs. Income and Dropoffs vs. Income; expectation would be that pickups and dropoffs would be highly correlated with income.

```{r}
#Prepares dataframe for the pickup rank and ranks those values
topPick<-df %>% dplyr::group_by(pickup_area) %>% tally(sort=TRUE)
topPick$pickrank<-rank(-topPick$n,ties.method="max")

#Prepares dataframe for the income rank and ranks those values
topInc<-df %>% dplyr::group_by(MedIncome,pickup_area) %>% tally(sort=TRUE)
topInc$incrank<-rank(-topInc$MedIncome)

#Joins the two datasets by their pickup_area names
topPick <- topPick %>% inner_join(topInc,incrank,by=c("pickup_area"))

#Drops a redundant column
topPick$n.y <- NULL

#Graphs these bad boys
ggplot(data=topPick) +
  geom_point(mapping=aes(x=incrank, y=pickrank, size=n.x)) +
  geom_abline(mapping = NULL, data = NULL, slope=1, intercept=0)

#We could do a top dropoff, but the MedIncome field is explicitly tied to pickup, which is probably fine 
#for what we're going after. IF we wanted it, as well, we could get it, but it might be a bit more 
#difficult than it's worth - it would require a new "MedIncome" column, that was tied to dropoffs


#topDO<-df %>% dplyr::group_by(pickup_area) %>% tally(sort=TRUE)
#topDO$droprank<-rank(-topDO$n,ties.method="max")

#topDO <- topDO %>% inner_join(topInc, incrank, by=c("pickup_area"))
#topDO$n.y <- NULL

#ggplot(data=topDO) +
#  geom_point(mapping=aes(x=incrank, y=droprank, size=n.x)) +
#  geom_abline(mapping = NULL, data = NULL, slope=1, intercept=0)
```

## Start Summary Tables

Create summary tables
* Ride count by Company
* Ride count by Dropoff Area
* Ride count by Pickup Area
* Rice count, avg fare, avg distance by pickup area
* Rice count, avg fare, avg distance by dropoffS area
```{r}
#Note: I am calling dplyr's functions directly to avoid issues with summarize
#Trips by Company
tibble.print_min=20
df %>% dplyr::group_by(company) %>% tally(sort=TRUE)
#Trips by Dropoff Area
topDrop <- df %>% dplyr::group_by(dropoff_area) %>% tally(sort=TRUE)
topDrop
#Trips by Origination and Destination
df %>% dplyr::group_by(pickup_area,dropoff_area) %>% tally(sort=TRUE)
#Trips by Pickup Area
topPickup <- df %>% dplyr::group_by(pickup_area) %>% tally(sort=TRUE)
topPickup
#Avg Fare and Distance by Pickup Area
df %>% dplyr::group_by(pickup_area) %>% dplyr::summarize(total_trips=n(),mean_fare = round(mean(fare,na.rm=TRUE),2),mean_distance=round(mean(trip_miles,na.rm=TRUE),1))
#Avg Fare and Distance by Dropoff Area
df %>% dplyr::group_by(dropoff_area) %>% dplyr::summarize(total_trips=n(),percent_trips=round(100*n()/nrow(df),5),
  median_fare = round(median(fare,na.rm=TRUE),2),
  mean_distance=round(mean(trip_miles,na.rm=TRUE),1)
  ) %>% arrange(desc(median_fare))
#Avg Fare and Distance by Pickup and Dropoff Areas
df %>% drop_na() %>% filter(fare<900) %>% dplyr::group_by(pickup_area,dropoff_area) %>% dplyr::summarize(total_trips=n(),
  mean_fare = round(mean(fare,na.rm=TRUE),2),
  median_fare = round(median(fare,na.rm=TRUE),2),
  mean_hourly_rate = round(mean(totalHourlyRate,na.rm=TRUE),2),
  mean_hourly_fare = round(mean(totalHourlyFare,na.rm=TRUE),2),
  mean_tip=round(mean(tips,na.rm=TRUE),1),
  mean_mph=round(mean(trip_miles*60*60/trip_seconds,na.rm=TRUE),1),
  min_fare=round(min(fare,na.rm=TRUE),2),
  max_fare=round(max(fare,na.rm=TRUE),2),
  mean_distance=round(mean(trip_miles,na.rm=TRUE),1),
  median_distance=round(median(trip_miles,na.rm=TRUE),1),
  min_distance=round(min(trip_miles,na.rm=TRUE),1),
  max_distance=round(max(trip_miles,na.rm=TRUE),1),
  
  )  %>% arrange(desc(total_trips))
```


Generate summary calculations by month/year and plot.

```{r}
fareByMonthYear <- df %>% drop_na(c("startDate","fare")) %>% filter(fare<1000) %>% 
  dplyr::group_by(month=floor_date(startDate, "month")) %>%
   filter(month!="2016-10-01") %>% summarize(totalFares=sum(fare))
ggplot(fareByMonthYear, aes(x=month, y=totalFares, group=1))+
  geom_point()+geom_line()+ggtitle("Total Cab Fares in Chicago by Month/Year")+xlab("Month/Year")+ylab("Fares Paid within Chicago ($)")
```

```{r}
#Drop all NA on-the-fly
fareByMonth <- df %>% drop_na(c("startDate","fare")) %>%  filter(year(endDate)!='2016') %>% filter(fare<1000) %>% 
  dplyr::group_by(month=endMonth)  %>% summarize(total=sum(fare))
ggplot(fareByMonth, aes(x=month, y=total, group=1))+
  geom_point(size=3)+geom_line(size=1.25)+ggtitle("Total Cab Fares in Chicago by Calendar Month")+xlab("Calendar Month")+ylab("Total Fare ($)")
```

```{r}
#Throw out 10/2016 because zero rides
ridesByMonthYear <- df %>% dplyr::group_by(month=floor_date(df$startDate, "month")) %>% 
   filter(month!="2016-10-01") %>% tally(sort=TRUE)
ggplot(ridesByMonthYear, aes(x=month, y=n, group=1))+
  geom_point()+geom_line()+ggtitle("Number of Cab Rides by Month/Year")+xlab("Month/Year")+ylab("Rides")
```

Let's look at the total number of cab rides by calendar month. Since we only have data for the first 9 months of 2016, we drop all 2016 data to avoid introducing a bias in the trend.

```{r}
#Drop all NA on-the-fly
ridesByMonth <- df %>% drop_na(c("endMonth")) %>%  filter(year(endDate)!='2016') %>% 
  dplyr::group_by(month=endMonth)  %>%summarize(total=sum(!is.na(endMonth)))
ggplot(ridesByMonth, aes(x=month, y=total, group=1))+
  geom_point(size=3)+geom_line(size=1.25)+ggtitle("Number of Cab Rides by Calendar Month")+xlab("Calendar Month")+ylab("Rides")
```
```{r}
#Drop all NA on-the-fly
ridesByWeekday<- df %>% drop_na(c("endDay")) %>% dplyr::group_by(weekday=endDay) %>%
   summarize(total=sum(!is.na(endDay)))
ggplot(ridesByWeekday, aes(x=weekday, y=total, group=1))+
  geom_point(size=4)+geom_line(size=1.25)+ggtitle("Number of Cab Rides by Weekday")+xlab("Weekday")+ylab("Rides")
```

```{r}
#Drop all NA on-the-fly
fareByWeekday<- df %>% drop_na(c("endDay","fare")) %>% dplyr::group_by(weekday=endDay) %>%  filter(fare<1000) %>% 
   summarize(total=sum(fare))
ggplot(fareByWeekday, aes(x=weekday, y=total, group=1))+
  geom_point(size=4)+geom_line(size=1.25)+ggtitle("Total Fare by Weekday")+xlab("Weekday")+ylab("Fare ($)")
```

```{r}
#Drop all NA on-the-fly
avgFareByWeekday<- df %>% drop_na(c("endDay","fare")) %>% dplyr::group_by(weekday=endDay) %>%  filter(fare<1000) %>% 
   summarize(mean=mean(fare))
ggplot(avgFareByWeekday, aes(x=weekday, y=mean, group=1))+
  geom_point(size=4)+geom_line(size=1.25)+ggtitle("Avg Fare by Weekday")+xlab("Weekday")+ylab("Fare ($)")
```

```{r}
#Drop all NA on-the-fly
avgDistanceByWeekday<- df %>% drop_na(c("endDay","trip_miles")) %>% dplyr::group_by(weekday=endDay) %>%  filter(trip_miles<1000) %>% 
   summarize(mean=mean(trip_miles))
ggplot(avgDistanceByWeekday, aes(x=weekday, y=mean, group=1))+
  geom_point(size=4)+geom_line(size=1.25)+ggtitle("Avg Trip Distance by Weekday")+xlab("Weekday")+ylab("Distance (miles)")
```

## Statistics by Taxi Medallion
```{r}
#Throw out 10/2016 because zero rides
taxisByMonth <- df %>% dplyr::group_by(month=floor_date(df$startDate, "month")) %>% 
   filter(month!="2016-10-01") %>% summarize(cabs=n_distinct(taxi_id))
ggplot(taxisByMonth, aes(x=month, y=cabs, group=1))+
  geom_point()+geom_line()+ggtitle("Number of Taxis Operating by Month/Year")+xlab("Month/Year")+ylab("Taxis")
```

```{r}
#Throw out 10/2016 because zero rides
fareStatsByTaxi <- df %>% dplyr::group_by(taxi=taxi_id) %>% filter(fare<500 & fare>=2.25) %>%
   summarize(AvgFare=mean(fare),MedianFare=median(fare),MaxFare=max(fare),Trips=n())%>% arrange(desc(Trips))
fareStatsByTaxi
ggplot(fareStatsByTaxi, aes(x=Trips, y=MedianFare, group=1))+
  geom_point()+ggtitle("Median Fare by Taxi as Function of Rides Given")+xlab("Rides Given")+ylab("Median Fare per Ride")

```
```{r}
#Throw out 10/2016 because zero rides
tipStatsByTaxi <- df %>% drop_na(c("tips")) %>% dplyr::group_by(taxi=taxi_id) %>% filter(tips>=0) %>%
   summarize(AvgTip=mean(tips),MedianTip=median(tips),MaxTip=max(tips),Trips=n()) %>% arrange(desc(Trips))
tipStatsByTaxi
ggplot(tipStatsByTaxi, aes(x=Trips, y=MedianTip, group=1))+
  geom_point()+ggtitle("Median Tip by Taxi as Function of Rides Given")+xlab("Rides Given")+ylab("Median Tip per Ride")
ggplot(tipStatsByTaxi, aes(x=Trips, y=AvgTip, group=1))+
  geom_point()+ggtitle("Avg Tip by Taxi as Function of Rides Given")+xlab("Rides Given")+ylab("Avg Tip per Ride")

```

tipStatsByTaxi <- df %>% drop_na(c("tips")) %>% dplyr::group_by(taxi=taxi_id) %>% filter(tips>=0) %>%
   summarize(AvgTip=mean(tips),MedianTip=median(tips),MaxTip=max(tips),Trips=n()) %>% arrange(desc(Trips))
tipStatsByTaxi
```{r}
tipStatsByPickup <- df %>% drop_na(c("tips","trip_seconds","pickup_area")) %>% dplyr::group_by(pickup_area) %>% 
   filter(n()>100) %>% filter(trip_seconds>0)  %>%
   summarize(AvgTip=round(mean(tips),2),
             AvgTipPerMin=round(mean(tips/(trip_seconds/60)),2),
             MedianTip=median(tips),
             MedianTipPerMin=round(median(tips/(trip_seconds/60)),2),
             MaxTip=max(tips),
             NoTip=sum(ifelse(tips==0,1,0)),
             PercentTip=round(100*(sum(ifelse(tips==0,0,1)))/n(),1),
             Trips=n()) %>% 
  arrange(desc(PercentTip))
tipStatsByPickup
```
