---
title: "Final Project"
subtitle: "Data Exploration"
author: ""
date: "2019-Feb-25"
fontsize: 8 pt
output:
  #    beamer_presentation: #everything after hash "#" will be ignored 
pdf_document:
number_sections: false
fig_width: 6
fig_height: 4
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      #include = TRUE, 
                      eval = TRUE, 
                      
                      fig.width = 6, fig.height = 4,
                      results='hold',
                      warning = FALSE,
                      message=FALSE,
                      cache = TRUE,
                      digits = 3) 
```

```{r,echo=FALSE,results='hide', message=FALSE}
# Packages and Data Parser
#Note: Be sure plyr is loaded before dplyr
PackageList =c("plyr","dplyr","tidyverse","lubridate","censusapi","forcats","igraph","network","sna","visNetwork","threejs","networkD3","ndtv")
NewPackages=PackageList[!(PackageList %in% 
                            installed.packages()[,"Package"])]
if(length(NewPackages)) install.packages(NewPackages,repos = "http://cran.us.r-project.org")
lapply(PackageList,require,character.only=TRUE)#array function

options(tibble.print_max = Inf, tibble.print_min = 20) 
set.seed(1) #Always set the seed for reproducibility
```

Load data and begin generating exploratory plots.

```{r}
source("taxiDatazw.R")
#Load records from taxi data source and store as a dataframe
data1m <- taxiDataZW(1000000)
df <- data.frame(data1m$rawData)
rm(data1m) #Drop old data
df$dropoff_community_area<-as.factor(df$dropoff_community_area)
df$dropoff_centroid_location.coordinates<-as.character(df$dropoff_centroid_location.coordinates)
df$pickup_centroid_location.coordinates<-as.character(df$pickup_centroid_location.coordinates)
df$trip_total<-as.numeric(df$trip_total)
#Load census data by Chicago Community Area
ccaData <- read.csv("ChicagoCCAData.csv")
ccaData$CCA.Code<-as.factor(ccaData$CCA.Code)
ccaData$GEOG<-as.factor(ccaData$GEOG)
#Look up name of pickup CCA and add to df
df<- df %>% left_join(select(ccaData,GEOG,CCA.Code),by=c("pickup_community_area"="CCA.Code"))
names(df)[names(df) == 'GEOG'] <- 'pickup_area'
#Look up name of dropoff CCA and add to df
df<- df %>% left_join(select(ccaData,GEOG,CCA.Code),by=c("dropoff_community_area"="CCA.Code"))
names(df)[names(df) == 'GEOG'] <- 'dropoff_area'
#Look up population>16yo of pickup CCA and add to df
df<- df %>% left_join(select(ccaData,POP_16OV,CCA.Code),by=c("pickup_community_area"="CCA.Code"))
names(df)[names(df) == 'POP_16OV'] <- 'Pop16Plus'
#Look up Labor Force of pickup CCA and add to df
df<- df %>% left_join(select(ccaData,IN_LBFRC,CCA.Code),by=c("pickup_community_area"="CCA.Code"))
names(df)[names(df) == 'IN_LBFRC'] <- 'IN_LBFRC'
#Look up Unemp of pickup CCA and add to df
df<- df %>% left_join(select(ccaData,UNEMP,CCA.Code),by=c("pickup_community_area"="CCA.Code"))
names(df)[names(df) == 'UNEMP'] <- 'UNEMP'
#Look up Unemp of pickup CCA and add to df
df<- df %>% left_join(select(ccaData,MEDINC,CCA.Code),by=c("pickup_community_area"="CCA.Code"))
names(df)[names(df) == 'MEDINC'] <- 'MedIncome'
```


Clean data. For instance, "Chicago Elite Cab Corp. (Chicago Carriag" and "Chicago Elite Cab Corp." are clearly the same firm. We replace the former with the latter. 
We add factors for the Start Date (as opposed to date time), End Date, and Start Day of Week.
* In company, set "Chicago Elite Cab Corp. (Chicago Carriag" equal to "Chicago Elite Cab Corp."
* Add factor for Start Date (not date time)
* Add factor for End Date (not date time)
* Add factor for End Day of week
* Add factor for End month
* Add factor for End hour

```{r}
df$company[df$company=="Chicago Elite Cab Corp. (Chicago Carriag"] <- "Chicago Elite Cab Corp."
df$startDate <- date(df$trip_start_timestamp)
df$endDate <- date(df$trip_end_timestamp)
df$endDay <- wday(df$trip_end_timestamp,label=TRUE)
df$endMonth <- month(df$trip_end_timestamp,label=TRUE)
df$endHour <- hour(str_replace(df$trip_end_timestamp,"T"," "))
```

Throw out all rows where NAs exist for dropoff/pickup location and convert NA to explicit for company (necessary in summary calcs)

```{r}
#Uncomment the following line to drop all records where dropoff_area and pickup_area NA
df<-df %>% drop_na(c("dropoff_area","pickup_area"))
#Replace <NA> with string "NA" which can be used for grouping
df$company <- fct_explicit_na(df$company,"NA")
df$pickup_area <- fct_explicit_na(df$pickup_area,"NA")
df$dropoff_area <- fct_explicit_na(df$dropoff_area,"NA")
```

## Start Summary Tables

Create summary tables
* Ride count by Company
* Ride count by Dropoff Area
* Ride count by Pickup Area
* Rice count, avg fare, avg distance by pickup area
* Rice count, avg fare, avg distance by dropoffS area
```{r}
#Note: I am calling dplyr's functions directly to avoid issues with summarize
#Trips by Company
df %>% dplyr::group_by(company) %>% tally(sort=TRUE)
#Trips by Dropoff Area
topDrop <- df %>% dplyr::group_by(dropoff_area) %>% tally(sort=TRUE)
topDrop
#Trips by Origination and Destination
df %>% dplyr::group_by(pickup_area,dropoff_area) %>% tally(sort=TRUE)
#Trips by Pickup Area
topPickup <- df %>% dplyr::group_by(pickup_area) %>% tally(sort=TRUE)
topPickup

#Avg Fare and Distance by Pickup and Dropoff Areas
df %>% dplyr::group_by(pickup_area,dropoff_area) %>% dplyr::summarize(total_trips=n(),
  total_fares = round(sum(fare,na.rm=TRUE),2),                                     total_tips = round(sum(tips,na.rm=TRUE) ,2),
  avg_tip=round(mean(tips,na.rm=TRUE),2),
  avg_tip_percentFare=round(100*mean(tips/fare,na.rm=TRUE),1),
  mean_fare = round(mean(fare,na.rm=TRUE),2),
  median_fare = round(median(fare,na.rm=TRUE),2),
  min_fare=round(min(fare,na.rm=TRUE),2),
  max_fare=round(max(fare,na.rm=TRUE),2),
  mean_distance=round(mean(trip_miles,na.rm=TRUE),1),
  median_distance=round(median(trip_miles,na.rm=TRUE),1),
  min_distance=round(min(trip_miles,na.rm=TRUE),1),
  max_distance=round(max(trip_miles,na.rm=TRUE),1)
  )  %>% arrange(desc(total_trips))
```

